Locality and Effect Reflection

### 2025-07-10 paper.tex
More edits

\documentclass[acmsmall, screen, nonacm]{acmart}

@@

\begin{abstract}
  Algebraic effects and handlers allow the programmer to extend their
  language's \emph{ambient monad} \textbf{--} the monad describing which
  effects ordinary functions can perform \textbf{--} with additional
  effectful operations.

  Locality captures the general idea of a value depending on some notion
  of ``location''. We can represent it in a type system using a
  comonadic \emph{global modality} $\glob$ that tracks which values do
  not depend on their location. A particularly useful notion of
  ``location'' is the ambient monad.

  We introduce an alternative presentation of algebraic effects and
  handlers, \emph{effect reflection}, that provides, for any algebraic
  signature $\Sigma$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}$ is an inductive type representing the free
  algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
  to inject terms of $\Sigma$ into the current ambient
  monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
  $\Sigma$. This presentation falls out naturally from a simple
  denotational semantics for algebraic effects in terms of presheaves
  over bounded Lawvere theories.

  Effect reflection lets you work with any algebraic monad using an
  ordinary function type, allowing effectful code to be written in
  direct style. This provides all the benefits of typed algebraic
  effects and handlers without requiring an effect system or other
  specific additions to the type system. We demonstrate effect
  reflection by implementing it as a library in
  OxCaml\cite{lorenzen2024oxidizing} and working through some classic
  examples of programming with effects.
\end{abstract}

@@

\subsection{Algebraic effects}
Algebraic effects and handlers are an increasingly popular approach to
programming with effects. Handlers allow the programmer to add
additional effects to the \emph{ambient monad} of their language -- the
monad describing which effects an ordinary function can perform -- for
some region of their program.

@@

Note that, in OCaml, the effects are \emph{untracked}: nothing prevents
attempting to perform an algebraic effect operation that is not
supported by the current ambient monad, resulting in an
\lstinline[style=oxcaml]{Unhandled} exception being raised.

Other languages track effects to ensure that operations are only
performed when they are supported. However, these require extending the
type system with explicit mechanisms for tracking effects, as in
Koka\cite{leijen2017type}, or for tracking a general notion of
capability, as in Effekt\cite{brachthauser2022effects}.

The ``Lightweight effect polymorphism'' approach of Brachthauser et
al. elegantly manages to track effects without requiring explicit
effect-tracking machinery in the type system. However, it is restricted
to second-class functions. We would like to extend this approach to
first-class functions without having to fall back to explicit
effect-tracking machinery.

\subsection{Locality and effect reflection}
Locality captures the general idea of a value depending on some notion
of ``location''. We can represent it in a type system using a comonadic
\emph{global modality} $\glob$ that tracks which values do not depend on
their location. A particularly useful notion of ``location'' is the
ambient monad.

Treating the ambient monad as a notion of location allows a language to
provide, for any algebraic signature $\Sigma$, an isomorphism:
\begin{equation*}
  \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
\end{equation*}
where $\ind{\Sigma}$ is an inductive type representing the free
algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
to inject terms of $\Sigma$ into the current ambient
monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
$\Sigma$. We call this isomorphism \emph{effect reflection}.

The ability to use the ordinary arrow type
$\yoneda{\Sigma} \mathbin{\rightarrow} \glob X$ in place of the monad
$\ind{\Sigma}(\glob X)$ allows us to conveniently write effectful code,
using multiple effects, in direct style, just as algebraic effects do.

\subsection{Effect reflection in OxCaml}
OxCaml\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=oxcaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=oxcaml]{[]t} and
\lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
implicitly global, we write
e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.

Using this support, with can provide effect reflection as a simple API
that we can use for programming with effects.  For example, we can use
it to write an effect handler equivalent to the OCaml one above:
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type 'a t =
      | Get : int t
      | Set : int -> unit t
  end

@@

Note that the locality of the handler value ensures that these
operations can only be performed when they are available in the ambient
monad. This provides the same guarantees as using an effect system
without any type system machinery beyond tracking locality.

\subsection{Denotational semantics} The design of effect reflection falls
out naturally from a simple denotational semantics of algebraic
effects. The underlying idea is to index the semantics by the ambient
monad. Rather than the semantics of a type being a set it is instead a
function from ambient monads to set: if you tell it what the ambient
monad is it can tell you what its set of values is.

The semantics is defined in terms of presheaves of \emph{bounded Lawvere
  theories}. Bounded Lawvere theories are a categorical representation
of algebraic theories: we represent the ambient algebraic monad by the
corresponding algebraic theory. The key constructs of effect reflection
are then defined as follows:

\begin{itemize}
\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebras. We use
  $\mon$ to define the semantics of computations.

\item The global modality $\glob$ on such presheaves is defined such
  that $\glob(P)(L) = P(\initial)$. In other words, for a given presheaf
  $P$ it gives the constant presheaf that always ignores the ambient
  monad and instead returns $P$ at $\initial$, where $\initial$ is the
  algebraic theory with no operations or equations. This allows us to
  restrict a type to its terms that do not depend on the ambient monad.

\item The semantics of our handler types $\yoneda{\Sigma}$ are given by
  the Yoneda embedding, and our effect reflection isomorphism becomes a
  simple consequence of the Yoneda lemma and some properties of the sums
  of algebraic theories.
\end{itemize}

@@

Values in OxCaml are associated with both a type -- which describe how
values can be introduced and eliminated -- and a \emph{mode} -- which
track properties related to operations other than introduction and
elimination. Modes are a product of a number of different axes, tracking
a variety of different properties, but here we are only interested in
one of them: \emph{locality}. Ignoring the other axes, a mode is either
\lstinline[style=oxcaml]{local} or \lstinline[style=oxcaml]{global}.

Locality tracks the ability of a value to leave \emph{regions}.  Regions
describe some period of time during a program's execution. Every
function call defines a corresponding region that, by default, lasts for
the entire extent of the call. This means that, by default,
\lstinline[style=oxcaml]{local} values cannot be returned from a
function.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value is local and would escape its region.} @*)
\end{lstlisting}

To support functions that return local values, the region of a function
can be ended early using the \lstinline[style=oxcaml]{exclave} construct.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

@@

  module Handler(O : Op) : sig
    type t
  end
\end{lstlisting}
For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
the type of mappings from the operations of \lstinline[style=oxcaml]{O}
into the operations of the current ambient monad. Alternatively,
\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

@@

This ability for higher-order functions like
\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by
Brachth{\"a}user et al.\cite{brachthauser2020effects} without requiring
a separate notion of second-class function.

@@

There is no fundamental reason that effect reflection should be
restricted to only one-shot continuations, so for the purposes of this
paper we shall ignore the restriction. However, for the same reasons as
OCaml, our actual implementation is restricted one-shot
continuations. Note that OxCaml's \lstinline[style=oxcaml]{once}
mode\cite{lorenzen2024oxidizing} allows us to enforce this restriction
statically.

@@

If we tried to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get and error:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_gen i (fun gh ->
      List.iter files (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        gen gh i)))
  (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
                       (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
  (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

@@

\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}. Terms in
the free algebras of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other\cite{hyland2006combining}. We can define
this type and associated isomorphism as:
\begin{lstlisting}[style=oxcaml]
  module Half_term (L : Op) (R : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
  end

@@

The literature on lexical effect handlers often ignores this issue. For
example, Xie et al. \cite{xie2020effect} proposed a system which only
supported directly nesting effect handlers and had no mechanism for
writing a composable handler for generators that could be moved between
different underlying asynchronous I/O handlers.

@@

@@

The design of effect reflection falls out naturally from a semantics in
terms of presheaves. To illustrate this semantics we first build a
simple language with effect reflection.

\subsection{The standard parts}

Our language is simply typed, featuring all the usual type-formers. It
uses a variation of fine-grained call-by-value: it has separate syntax
and typing judgements for values and computations. The syntax and typing
rules for these standard parts of the language are shown in
Fig.\ref{fig:typing-standard}, and there isn't much interesting to say
about them.
\begin{figure}
\begin{align*}
  \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
                              \mid \tau \rightarrow \tau \mid \ldots \\
  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
  \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \mid \ldots \\
  \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
                                 \mid \app{v}{v} \mid \\
                            & \case{v}{x}{c}{x}{c} \mid \ldots
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\lbind{\Gamma}{x}{\tau} \turnv x \types \tau}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \\ \Gamma \turnv v_1 \types \tau_1}
{\Gamma \turnv (v_1, v_2) \types \tau_1 \times \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \fst{v} \types \tau_1}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \snd{v} \types \tau_2}
\and
\inferrule{\\}
{\Gamma \turnv () \types 1}
\and
\inferrule{\Gamma \turnv v \types \tau_1}
{\Gamma \turnv \inl{v} \types \tau_1 + \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_2}
{\Gamma \turnv \inr{v} \types \tau_1 + \tau_2}
\and
\inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau}
{\Gamma \turnc \return{v} \types \tau}
\and
\inferrule{\Gamma \turnc c_1 \types \tau_1 \\
           \lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}
{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2 \\
           \Gamma \turnv v_2 \types \tau_1}
{\Gamma \turnc v_1 v_2 \types \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 + \tau_2 \\
           \lbind{\Gamma}{x_1}{\tau_1} \turnc c_1 \types \tau_3 \\
           \lbind{\Gamma}{x_2}{\tau_2} \turnc c_2 \types \tau_3}
{\Gamma \turnc \case{v}{x_1}{c_1}{x_2}{c_2} \types \tau_3}
\end{mathpar}
\caption{Standard syntax and typing rules}
\label{fig:typing-standard}
\end{figure}

@@

We track locality in the language via a global comonadic modality
$\glob$. In our setting being global means not depending on the ambient
monad. For example, the function type $1 \rightarrow 1$ represents
computations in the ambient monad -- e.g. if the ambient monad includes
state then such a function may mutate that state -- whereas the global
function type $\glob (1 \rightarrow 1)$ has only a single value: the
function that does nothing and returns $()$.

@@

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$ \cite{kiselyov2015freer}.

As with other algebraic effects systems in the literature, we only
support handlers for algebraic theories without any equations, even
though our semantics is defined in terms of arbitrary theories. We
define algebraic signatures $\Sigma$ as finite sets of operations:
\begin{equation*}
\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
\end{equation*}
An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
parameterized by $\tau_p$ and with arity $\tau_a$.

@@

The language provides a type $\yoneda{\Sigma}$ for each $\Sigma$ to
represent mappings from the operations of $\Sigma$ into the current
ambient monad. Effect reflection appears in the language as two
operations: reflection and reification. Reflection
$\reflectraw{\Sigma}{op}$ uses a value of type $\yoneda{\Sigma}$ to map
the operation $op$ from $\Sigma$ into the ambient monad.  Reification
$\reifyraw{\Sigma}$ extends the ambient monad with the operations from
$\Sigma$, providing a value of type $\yoneda{\Sigma}$ for constructing
those operations via reflection. Their syntax and typing rules are shown
in Fig.\ref{fig:typing-reflection}.

@@

Note that the reduction relations do not include $\eta$ expansion. They
also do not include some conversions you might expect for computations:
\begin{itemize}
\item There is no rule to convert
  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
\end{itemize}
These rules are excluded because they are not $\beta$ rules. Instead
they are better thought of as $\nu$ rules \cite{allais2013new}:
additional equalities between neutral terms that are immediate
consequences of structural induction. Rather than be included as part of
ordinary reduction they can be done as an additional standardization
pass on neutral terms after reduction and $\eta$-expansion have
completed. This staged approach to computing equality of terms works
because $\nu$ rules cannot produce additional $\beta$ redexes.

@@

The design of effect reflection falls out naturally from a simple
denotational semantics built on presheaves. The key idea is to index the
meaning of types by an ambient algebraic monad of effects: rather than
have the semantics of each type be a set of elements, the semantics of
each type is a mapping from the ambient monad to a set of elements.

@@

Hyland et al.\cite{hyland2006combining} investigate combining algebraic
effects, including the key property of sums of algebraic theories (our
Lemma \ref{lemma:sums}) on which handlers are based.

>>>>

\documentclass[acmsmall, screen, review, anonymous]{acmart}

@@

\begin{abstract}
  Programming languages increasingly support algebraic effects and
  handlers for writing effectful code in a principled way by extending a
  language's \emph{ambient monad} \textbf{--} the monad describing which
  effects ordinary functions can perform \textbf{--} with additional
  effectful operations. However, integrating effects into type systems
  typically requires specialized effect systems that add complexity to
  the language.

  We introduce \emph{effect reflection}, an alternative presentation of
  algebraic effects and handlers that provides, for any algebraic
  signature $\Sigma$ and type $\tau$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau)
    \qquad \cong \qquad
    \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
  $\glob \tau$ \textbf{--} sometimes called the \emph{freer monad} of
  $\Sigma$, $\yoneda{\Sigma}$ is a type representing how to inject terms
  of $\Sigma$ into the current ambient monad, and $\glob$ is a comonadic
  \emph{global modality} used to track which values do not depend on the
  ambient monad. This isomorphism emerges naturally from a denotational
  semantics for algebraic effects in terms of presheaves over bounded
  Lawvere theories.

  Effect reflection provides all the expressiveness of typed algebraic
  effects without requiring language-level support for effect systems. We
  demonstrate this by implementing effect reflection as a library in
  OxCaml~\cite{lorenzen2024oxidizing} and showing how classic effectful
  programs can be written in natural direct style.
\end{abstract}

@@

\subsection{Algebraic effects}
Algebraic effects and handlers have emerged as a powerful approach to
structured programming with effects. Handlers allow the programmer to
temporarily extend the \emph{ambient monad} of their language -- the
monad describing which effects an ordinary function can perform -- with
additional effectful operations.

@@

However, effect handlers face a fundamental tension: \emph{untracked}
effects are unsafe, while \emph{tracked} effects require complex type
system extensions.

In OCaml, the effects are untracked: nothing prevents attempting to
perform an algebraic effect operation that is not supported by the
current ambient monad, resulting in an
\lstinline[style=oxcaml]{Unhandled} exception being raised.

In languages like Koka~\cite{leijen2017type} the type system is extended
to explicitly track the effects that a function may perform. This
prevents the unsafety on an unhandled effect but adds complexity to the
language.

The ``effects as capabilities'' approach of the Effekt
language~\cite{brachthauser2020effects} points the way by elegantly
tracking effects without requiring explicit effect-tracking machinery in
the type system. However, it is restricted to second-class
functions\footnote{That restriction is lifted in more recent work on Effekt~\cite{brachthauser2022effects}, but only by falling back to explicit
effect-tracking machinery.}. We aim to extend this approach to first
class functions.

\subsection{Locality and effect reflection}
Locality captures the general idea of a value depending on some notion
of ``location''. We can represent it in a type system using a comonadic
\emph{global modality} $\glob$ that tracks which values do not depend on
their location. Our key insight is that the ambient monad can be viewed
as a form of location and controlled using $\glob$.

Based on this insight, we present \emph{effect reflection}, providing,
for any algebraic signature $\Sigma$ and type $\tau$, an isomorphism:
\begin{equation*}
  \ind{\Sigma}(\glob \tau)
  \qquad \cong \qquad
  \yoneda{\Sigma} \rightarrow \glob \tau
\end{equation*}
where $\ind{\Sigma}(\glob \tau)$ is an inductive type representing the
free algebra of $\Sigma$ at $\glob \tau$ \textbf{--} sometimes called
the \emph{freer monad} of $\Sigma$, and $\yoneda{\Sigma}$ is a type
representing how to inject terms of $\Sigma$ into the current ambient
monad.

This isomorphism reveals that effectful computations can be represented
as ordinary functions, enabling direct-style programming with full
effect safety.

\subsection{Effect reflection in OxCaml}
OxCaml~\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=oxcaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=oxcaml]{[]t} and
\lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
implicitly global we write \lstinline[style=oxcaml]{@ local} on them,
e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.

Using this support, we can provide effect reflection as a simple API.
We represent algebraic signatures as modules with a single type
constructor, providing $\ind{\Sigma}$ as
\lstinline[style=oxcaml]{'a Term(Sigma).t} and $\yoneda{Sigma}$ as
\lstinline[style=oxcaml]{Handler(Sigma).t}.

For example, we can use it to write an effect handler equivalent to the
OCaml one above:
\begin{lstlisting}[style=oxcaml]
  module State = struct
    type 'a t =
      | Get : int t
      | Set : int -> unit t
  end

@@

The locality of the \lstinline[style=oxcaml]{Handler(State).t} value
passed to \lstinline[style=oxcaml]{f} ensures that these operations can
only be performed when available in the ambient monad. The type system
prevents this handler value from escaping the scope of
\lstinline[style=oxcaml]{f}, which in turn prevents
\lstinline[style=oxcaml]{get} or \lstinline[style=oxcaml]{set} from
being called outside that scope.

\subsection{Denotational semantics}

This design emerges naturally from a denotational semantics where types
are interpreted as presheaves over \emph{bounded Lawvere theories} -- a
categorical framework that makes the connection between locality and
effects precise.

The key idea is to index the semantics by the ambient monad: rather than
interpreting types as sets, we interpret them as functions from ambient
monads to sets. We represent ambient monads using their corresponding
algebraic theories in the framework of bounded Lawvere theories.

The key constructs of effect reflection are then defined as follows:
\begin{itemize}
\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebras.

\item The global modality $\glob P$ is defined as the constant presheaf
  that always ignores the ambient monad and instead returns
  $P(\initial)$, where $\initial$ is the algebraic theory with no
  operations or equations.

\item The semantics of our handler types $\yoneda{\Sigma}$ are given by
  the Yoneda embedding

\item The effect reflection isomorphism becomes a simple consequence of
  the Yoneda lemma and some properties of the sums of algebraic
  theories.
\end{itemize}

@@

Values in OxCaml are associated with both a type -- which describes how
values can be introduced and eliminated -- and a \emph{mode} -- which
tracks properties related to operations other than introduction and
elimination. Modes are a product of a number of different axes, tracking
a variety of different properties, but here we are only interested in
one of them: \emph{locality}. Ignoring the other axes, a mode is either
\lstinline[style=oxcaml]{local} or \lstinline[style=oxcaml]{global}.

Locality tracks whether a value can 'escape' the current
\emph{region}. By default, the current region ends when a function
returns, so local values cannot be returned from functions.
\begin{lstlisting}[style=oxcaml]
  let leak_local x =
    let (o @ local) = Some x in
    o
  (*@ \textcolor{red}{Error: o} @*)
         (*@ \textcolor{red}{\textasciicircum} @*)
  (*@ \textcolor{red}{This value is local and would escape its region.} @*)
\end{lstlisting}

The \lstinline[style=oxcaml]{exclave} construct allows functions to
return local values.
\begin{lstlisting}[style=oxcaml]
  let return_local x =
    exclave
      (let (o @ local) = Some x in
       o)
\end{lstlisting}
\lstinline[style=oxcaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

@@

  module Handler(O : Op) : sig
    type t
  end
\end{lstlisting}
These types directly implement the types from our core isomorphism:
\lstinline[style=oxcaml]{'a Term(Sigma).t} corresponds to
$\ind{\Sigma}(\glob \tau)$, while
\lstinline[style=oxcaml]{Handler(Sigma).t} corresponds to
$\yoneda{\Sigma}$

For an effect signature \lstinline[style=oxcaml]{O},
\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
algebra of the corresponding effect -- otherwise known as the freer
monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
the type of mappings from the operations of \lstinline[style=oxcaml]{O}
into the operations of the current ambient monad. Alternatively,
\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
algebraic effect handlers that are handling the current computation --
hence the name.

@@

This ability for higher-order functions like
\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
monads achieves the ``lightweight effect polymorphism'' described by
Brachth{\"a}user et al.~\cite{brachthauser2020effects} without requiring
second-class functions -- locality provides the necessary restrictions
while maintaining first-class status.

@@

There is no fundamental reason that effect reflection should be
restricted to only one-shot continuations, so for the purposes of this
paper we shall ignore the restriction. However, for the same reasons as
OCaml, our actual implementation is restricted one-shot
continuations. Note that OxCaml's \lstinline[style=oxcaml]{once}
mode~\cite{lorenzen2024oxidizing} allows us to enforce this restriction
statically.

@@

If we tried to do something similar using our
\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
\begin{lstlisting}[style=oxcaml]
  handle_await (fun ch ->
    handle_gen i (fun gh ->
      List.iter files (fun fl ->
        let s = await ch (File.read fl) in
        let i = int_of_string s in
        gen gh i)))
  (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
                       (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
  (*@ \textcolor{red}{ This value is local and would escape its region.} @*)
\end{lstlisting}

@@

\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}. Terms in
the free algebras of the sum of two effects
(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
recursive datatype that is built from the terms of one of the effects
and the operations of the other~\cite{hyland2006combining}. We can define
this type and associated isomorphism as:
\begin{lstlisting}[style=oxcaml]
  module Half_term (L : Op) (R : Op) : sig
    type 'a t =
      | Return : 'a @@ global -> 'a t
      | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
  end

@@

The literature on lexical effect handlers often ignores this issue. For
example, Xie et al.~\cite{xie2020effect} proposed a system which only
supported directly nesting effect handlers and had no mechanism for
writing a composable handler for generators that could be moved between
different underlying asynchronous I/O handlers.

@@

\subsection{Summary}

This section has demonstrated how effect reflection provides a practical
and expressive approach to algebraic effects through a simple library
API. The key benefits include: safety guarantees through locality
without requiring specialized effect systems, seamless composition of
effects through nesting for common cases, support for more complex
composition patterns through effect forwarding when needed, and the
ability to achieve lightweight effect polymorphism with first-class
functions. The examples show that effect reflection can handle the full
spectrum of algebraic effects—from simple state management to complex
asynchronous I/O and generators—while maintaining the direct-style
programming that makes algebraic effects appealing. Most importantly,
the approach scales from basic usage to sophisticated effect composition
without requiring programmers to learn complex type system machinery
beyond locality tracking.

@@

The design of effect reflection falls out naturally from a denotational
semantics in terms of presheaves. To make this connection precise and
demonstrate the key properties of effect reflection, we present a
simply-typed calculus that directly incorporates the constructs needed
for effect reflection.

\subsection{The base language}

Our language is simply typed, featuring all the usual type-formers. It
uses a variation of fine-grained call-by-value: it has separate syntax
and typing judgements for values and computations. The syntax and typing
rules for the base language are shown in Fig.\ref{fig:typing-standard}.
\begin{figure}
\begin{align*}
  \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
                              \mid \tau \rightarrow \tau \mid \ldots \\
  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
  \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \mid \ldots \\
  \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
                                 \mid \app{v}{v} \mid \\
                            & \case{v}{x}{c}{x}{c} \mid \ldots
\end{align*}
\begin{mathpar}
\inferrule{\\}
{\lbind{\Gamma}{x}{\tau} \turnv x \types \tau}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \\ \Gamma \turnv v_1 \types \tau_1}
{\Gamma \turnv (v_1, v_2) \types \tau_1 \times \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \fst{v} \types \tau_1}
\and
\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
{\Gamma \turnv \snd{v} \types \tau_2}
\and
\inferrule{\\}
{\Gamma \turnv () \types 1}
\and
\inferrule{\Gamma \turnv v \types \tau_1}
{\Gamma \turnv \inl{v} \types \tau_1 + \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_2}
{\Gamma \turnv \inr{v} \types \tau_1 + \tau_2}
\and
\inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau}
{\Gamma \turnc \return{v} \types \tau}
\and
\inferrule{\Gamma \turnc c_1 \types \tau_1 \\
           \lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}
{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2}
\and
\inferrule{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2 \\
           \Gamma \turnv v_2 \types \tau_1}
{\Gamma \turnc v_1 v_2 \types \tau_2}
\and
\inferrule{\Gamma \turnv v \types \tau_1 + \tau_2 \\
           \lbind{\Gamma}{x_1}{\tau_1} \turnc c_1 \types \tau_3 \\
           \lbind{\Gamma}{x_2}{\tau_2} \turnc c_2 \types \tau_3}
{\Gamma \turnc \case{v}{x_1}{c_1}{x_2}{c_2} \types \tau_3}
\end{mathpar}
\caption{Standard syntax and typing rules}
\label{fig:typing-standard}
\end{figure}

@@

We track locality in the language via a global comonadic modality
$\glob$. In our setting being global means not depending on the ambient
monad. For example, $1 \rightarrow 1$ represents computations that may
use operations from the ambient monad, while $\glob (1 \rightarrow 1)$
contains only pure computations that ignore the ambient monad.

@@

Effect reflection requires, for any signature $\Sigma$ that is being
reflected, an inductive type $\ind{\Sigma}$ to represent the free
algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
  monad} of $\Sigma$~\cite{kiselyov2015freer}.

As is standard for algebraic effects systems, we only support handlers
for algebraic theories without any equations, even though our semantics
is defined in terms of arbitrary theories. We define algebraic
signatures $\Sigma$ as finite sets of operations:
\begin{equation*}
\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
\end{equation*}
An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
parameterized by $\tau_p$ and with arity $\tau_a$.

@@

The language provides a type $\yoneda{\Sigma}$ for each $\Sigma$ to
represent mappings from the operations of $\Sigma$ into the current
ambient monad. Effect reflection appears in the language as two
operations: reflection and reification. Reflection
$\reflectraw{\Sigma}{op}$ uses a value of type $\yoneda{\Sigma}$ to map
the operation $op$ from $\Sigma$ into the ambient monad.  Reification
$\reifyraw{\Sigma}$ extends the ambient monad with the operations from
$\Sigma$, providing a value of type $\yoneda{\Sigma}$ for constructing
those operations via reflection.  These operations implement the core
isomorphism of effect reflection: reification can convert functions
$\yoneda{\Sigma} \rightarrow \glob \tau$ into terms $\ind{\Sigma}(\glob
\tau)$, while reflection provides the inverse direction. Their syntax
and typing rules are shown in Fig.\ref{fig:typing-reflection}.

@@

Note that the reduction relations do not include $\eta$ expansion. They
also do not include some conversions you might expect for computations:
\begin{itemize}
\item There is no rule to convert
  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
\end{itemize}
These rules are excluded because they are not $\beta$ rules. Instead
they are better thought of as $\nu$ rules~\cite{allais2013new}:
additional equalities between neutral terms that are immediate
consequences of structural induction. Rather than be included as part of
ordinary reduction they can be done as an additional standardization
pass on neutral terms after reduction and $\eta$-expansion have
completed. This staged approach to computing equality of terms works
because $\nu$ rules cannot produce additional $\beta$ redexes.

@@

The design of effect reflection falls out naturally from a simple
denotational semantics built on presheaves. This semantic foundation not
only validates our approach but also makes the connection between
locality and effects precise. The key idea is to index the meaning of
types by an ambient algebraic monad of effects: rather than have the
semantics of each type be a set of elements, the semantics of each type
is a mapping from the ambient monad to a set of elements.

@@

Hyland et al.~\cite{hyland2006combining} investigate combining algebraic
effects, including the key property of sums of algebraic theories (our
Lemma \ref{lemma:sums}) on which handlers are based.
